---
title: "Conference Tier list"
output:
  pdf_document: default
  html_document: default
date: "2025-05-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#
# PROJECT: The Conference Premium
# GOAL: Rank Conferences by their 'Draft Boost'
# MODEL: Multivariate Regression (Draft Pick ~ Stats + Conference)
#

library(tidyverse)

# 1. LOAD DATA
col_stats <- read.csv("college_statistics.csv")
draft     <- read.csv("nfl_draft_prospects.csv")

# 2. DEFINE THE CONFERENCES (Manual Mapping)
# We map the major schools to their Power 5 Conference
get_conference <- function(school) {
  if (school %in% c("Alabama", "LSU", "Georgia", "Florida", "Auburn", "Texas A&M", 
                    "Ole Miss", "Mississippi State", "Tennessee", "Arkansas", 
                    "Kentucky", "Missouri", "South Carolina", "Vanderbilt")) {
    return("SEC")
  } else if (school %in% c("Ohio State", "Michigan", "Penn State", "Wisconsin", 
                           "Iowa", "Michigan State", "Minnesota", "Northwestern", 
                           "Indiana", "Purdue", "Illinois", "Maryland", "Rutgers", "Nebraska")) {
    return("Big Ten")
  } else if (school %in% c("Clemson", "Miami (FL)", "Florida State", "North Carolina", 
                           "Virginia Tech", "Louisville", "NC State", "Pittsburgh", 
                           "Virginia", "Wake Forest", "Georgia Tech", "Boston College", 
                           "Syracuse", "Duke", "Notre Dame")) {
    return("ACC")
  } else if (school %in% c("Oklahoma", "Texas", "Oklahoma State", "TCU", "Baylor", 
                           "Iowa State", "Kansas State", "West Virginia", "Texas Tech", "Kansas")) {
    return("Big 12")
  } else if (school %in% c("USC", "Oregon", "Washington", "Utah", "Stanford", 
                           "Arizona State", "California", "UCLA", "Washington State", 
                           "Arizona", "Oregon State", "Colorado")) {
    return("Pac-12")
  } else {
    return("Group of 5 / FCS")
  }
}

# 3. PREPARE COLLEGE STATS (WRs)
stats_clean <- col_stats %>%
  filter(pos_abbr == "WR") %>%
  filter(statistic %in% c("Receiving Yards", "Receiving Touchdowns")) %>%
  group_by(player_name, school) %>%
  summarize(
    Total_Yards = sum(value, na.rm = TRUE),
    Total_TDs   = sum(value, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(Total_Yards > 1000) # Only look at productive players

# 4. PREPARE DRAFT DATA
draft_clean <- draft %>%
  filter(position == "Wide Receiver") %>%
  select(player_name, Draft_Pick = overall, Height = height, Weight = weight) %>%
  drop_na(Draft_Pick)

# 5. MERGE & APPLY CONFERENCE MAPPING
conf_data <- inner_join(stats_clean, draft_clean, by = "player_name") %>%
  rowwise() %>%
  mutate(Conference = get_conference(school)) %>%
  ungroup()

# Set "Group of 5 / FCS" as the Baseline for comparison
conf_data$Conference <- relevel(as.factor(conf_data$Conference), ref = "Group of 5 / FCS")

print(paste("Players Analyzed:", nrow(conf_data)))

# 6. THE MODEL: RANKING THE CONFERENCES
# The coefficients tell us how many picks EARLIER a player goes compared to a G5 player
conf_model <- lm(Draft_Pick ~ Total_Yards + Total_TDs + Conference, data = conf_data)

summary(conf_model)

# 7. VISUALIZATION: THE CONFERENCE TIER LIST
# We plot the Average Draft Pick by Conference
avg_picks <- conf_data %>%
  group_by(Conference) %>%
  summarize(Avg_Pick = mean(Draft_Pick), Count = n()) %>%
  arrange(Avg_Pick) # Lower is Better

ggplot(avg_picks, aes(x = reorder(Conference, Avg_Pick), y = Avg_Pick, fill = Conference)) +
  geom_bar(stat = "identity", color = "black") +
  coord_flip() + # Horizontal Bars
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "The Conference Premium: Average Draft Position by League",
    subtitle = "Lower Bar = Higher Draft Capital (Better)",
    x = "Conference",
    y = "Average Draft Pick Number"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Introduction** Draft capital is a front office's most valuable asset. This project mathematically investigates the implicit "Draft Boost" awarded to collegiate wide receivers strictly based on their conference affiliation, aiming to determine if teams overvalue players from major conferences when production is held constant.

**Methodology & Data**

-   **Data Engineering & Filtering:** I ingested college statistics and NFL draft prospect data, mapping schools to their respective Power 5 or Group of 5 conferences. To eliminate noise from non-prospects, the dataset was strictly filtered to include only wide receivers with over 1,000 career collegiate receiving yards.

-   **Modeling (Multivariate Linear Regression):** I built an lm model setting the draft pick as the target variable. The model explicitly controlled for on-field production (total yards and total tds) while evaluating the categorical Conference variable.

-   **Statistical Baseline:** I releveled the factor to set "Group of 5 / FCS" as the mathematical baseline. This allowed the model's coefficients to output the exact number of picks earlier a player is drafted simply for wearing a Power 5 helmet.

**Summary & Outcomes** By isolating conference affiliation from raw production, the regression model successfully quantifies regional scouting biases. The resulting visualization aggregates the average draft pick by conference, providing actionable intelligence for exploiting value in later rounds by targeting highly productive players from smaller programs that the market systematically undervalues.
